Typed Lua: An Optional Type System for Lua

### 2014-03-14 doc/paper/typedlua.tex
remove nil in map type
rewrite of section on objects and classes

The concrete syntax of Typed Lua has syntax for
common table types. One syntax defines table types for
maps: it is written \texttt{\{ t: u \}},
and maps to the table type $\{t:u\}$.
This table type represents a map that maps values of type
$t$ to values of type $u \cup \Nil$.
Another syntax defines table types for arrays:
it is written \texttt{\{ t \}}, and maps to the table type
$\{\Number:t\}$. A third syntax sugar defines
table types for records:
it is written \texttt{\{ s1: t1, ..., sn: tn \}}, where
each $s_i$ is a literal number, string, or boolean, 
and maps to the table type $\{s_{1}:t_{1}, ..., s_{n}:t_{n}\}$,
where each $s_i$ is the corresponding literal type.

@@

Besides interfaces, Typed Lua will also introduce classes.
We want to offer classes as a standard way for writing code that
Lua programmers are already using, but without tying them to a
specific model.
Our classes will not include static attributes and will also not offer
privacy rules.

>>>>

The concrete syntax of Typed Lua has syntax for
common table types. One syntax defines table types for
maps: it is written \texttt{\{ t: u \}},
and maps to the table type $\{t:u\}$.
This table type represents a map that maps values of type
$t$ to values of type $u$.
Another syntax defines table types for arrays:
it is written \texttt{\{ t \}}, and maps to the table type
$\{\Number:t\}$. A third syntax sugar defines
table types for records:
it is written \texttt{\{ s1: t1, ..., sn: tn \}}, where
each $s_i$ is a literal number, string, or boolean, 
and maps to the table type $\{s_{1}:t_{1}, ..., s_{n}:t_{n}\}$,
where each $s_i$ is the corresponding literal type.

@@

We can use recursive table types to represent objects,
and Typed Lua has syntactic sugar to make defining these
types easier:

\begin{verbatim}
    interface Shape
      x, y: number
      const move: (dx: number, dy: number) => ()
    end
\end{verbatim}

The double arrow in the type of the two methods is
syntactic sugar for having a first parameter named
{\tt self} with the same type as the interface. The
{\tt const} qualifier is necessary for covariance
in the types of the methods, and to make subtyping
among object types work.

Type-checking a method call is straightforward, but
there is still the matter of how to construct a value with
the object type above. The following example shows one
way:

\begin{verbatim}
    local shape = { x = 0, y = 0 }
    const function shape:move(dx: number, 
                              dy: number)
      self.x = self.x + dx
      self.y = self.y + dy
    end
\end{verbatim}

The {\tt :} syntactic sugar that the example uses also comes
from Lua, and assigns a function to the field with a first
parameter named {\tt self}, plus any other parameters.

Lua has a mechanism for Self-like delegation of missing
fields in a table. After we do {\tt setmetatable(t1,
{ \_\_index = t2 })}, Lua looks up in {\tt t2} any missing
fields of {\tt t1}. Lua programmers often use this mechanism
to simulate classes, as in the following example:

\begin{verbatim}
    local Shape = { x = 0, y = 0 }
    const function Shape:new(x: number, y: number)
      local s = setmetatable({},
                             { __index = self })
      s.x = x
      s.y = y
      return shape
    end
    const function Shape:move(dx: number,
                              dy: number)
      self.x = self.x + dx
      self.y = self.y + dy
    end
    local shape1 = Shape:new(0, 5)
    local shape2 = Shape:new(10, 10)
\end{verbatim}

Typed Lua assigns the type
$\mathbf{Shape} \times \Number \times \Number \times
\Top* \rightarrow \mathbf{Shape} \times \Nil*$ to
{\tt new}. In a {\tt setmetatable} expression,
if the type $t_1$ of the first operand is open and a supertype
of the type $t_2$ of the second operand's {\_\_index} field,
it changes the type of the first operand to $t_1$, and
it remains open.

As the result of {\tt setmetatable} can be an open
table type, we can simulate single inheritance, and
override methods:

\begin{verbatim}
    local Circle = setmetatable({},
                                { __index = Shape })
    Circle.radius = 0
    const function Circle:new(x: number,
                              y: number,
                              radius: number)
      local c = setmetatable(Shape:new(x, y),
                             { __index = self })
      c.radius = radius
      return c
    end
    const function Circle:area()
      return math.pi * self.radius * self.radius
    end
\end{verbatim}

Notice how we can call {\tt Shape}'s constructor inside
the overriden constructor. A limitation of this class
system is that the overriden constructors must be a
subtype of the original constructor.

If we erase all type and {\tt const} annotations, the
two examples above are valid Lua code, with the same
semantics as the Typed Lua Code.

The current version of Typed Lua does not have a polymorphic
type system, so programmers currently cannot hide the
calls to {\tt setmetatable} behind nicer abstractions, as
some Lua libraries do. A few Lua programs also use
other features of {\tt setmetatable} which are currently
not typeable, such as operator overloading.
